---
title: "Running NCA with aNCA: R Script Walkthrough"
description: >
  Step-by-step guide to running NCA analyses with aNCA via R scripts, YAML settings, or the Shiny app.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running NCA with aNCA: R Script Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Introduction
The aNCA Shiny app exports a **settings YAML** and an **R script** that together
reproduce the analysis outside the app. This vignette walks through that R
script step by step, using the built-in `adnca_example` dataset with concrete
values for every setting.

By the end you will have:
- A PKNCAdata object with mapped columns, filters and intervals
- NCA results with flag rules, interval parameters and ratio calculations applied
- CDISC-compliant PP, ADPP, and ADNCA datasets
- A pivoted results table ready for review

## Start here: choose your workflow
Use this section to decide how you want to run NCA, based on your background and preferred workflow.

| If you are... | Best path | What you get |
|---|---|---|
| Comfortable running scripted analyses in R | [Path A: run the full script directly](#path-a-run-the-full-script-directly) | Maximum control and visibility over each step |
| Working from settings YAML files (e.g., from app exports or versioned config files) | [Path B: generate the same script from YAML](#path-b-generate-the-same-script-from-yaml-with-get_settings_code) | Reproducible script generation with `get_settings_code()` |
| Prefer playing with an App and an interface | [Path C: generate from the app](#path-c-generate-from-the-anca-app) | Build settings interactively, then export YAML and script |

## Install aNCA
Install the package from GitHub (the CRAN version still does not have the features shown here):

```{r install-dev-version, eval = FALSE}
# Option 1
remotes::install_github("pharmaverse/aNCA")

# Option 2
pak::pak("pharmaverse/aNCA")
```

Then restart your R session and confirm the installed version:

```{r check-version, eval = FALSE}
packageVersion("aNCA")
```

## Path A: run the full script directly

### 1. Load the package and data
```{r load, eval = FALSE}
library(aNCA)
library(dplyr)
# The example dataset ships with the package.
# In practice this would be: adnca_data <- read_pk("path/to/your/data.csv")
adnca_data <- adnca_example
head(adnca_data)
```
`adnca_example` is a 76-row ADNCA dataset with two analytes (DrugA and its
metabolite Metab-DrugA), two specimen types (SERUM and URINE), and two dosing
periods (DOSE 1, DOSE 2). Subjects received different dose levels via
EXTRAVASCULAR and INTRAVASCULAR routes.

### 2. Define the column mapping
The mapping tells aNCA which columns in your dataset correspond to the expected
CDISC variables. Keys match the standard ADNCA column names; values are the
actual column names in your data.
```{r mapping, eval = FALSE}
mapping <- list(
  STUDYID = "STUDYID",
  USUBJID = "USUBJID",
  DOSEA   = "DOSEA",
  DOSEU   = "DOSEU",
  DOSETRT = "DOSETRT",
  PARAM   = "PARAM",
  Metabolites = "Metab-DrugA",
  ARRLT   = "ARRLT",
  NRRLT   = "NRRLT",
  AFRLT   = "AFRLT",
  NCAwXRS = c("NCA1XRS", "NCA2XRS"),
  NFRLT   = "NFRLT",
  PCSPEC  = "PCSPEC",
  ROUTE   = "ROUTE",
  TRTRINT = "TRTRINT",
  ADOSEDUR = "ADOSEDUR",
  Grouping_Variables = c("TRT01A", "RACE", "SEX"),
  RRLTU   = "RRLTU",
  VOLUME  = "VOLUME",
  VOLUMEU = "VOLUMEU",
  AVAL    = "AVAL",
  AVALU   = "AVALU",
  ATPTREF = "ATPTREF"
)
```
A few notes:
- **`Metabolites`**: The PARAM value(s) that represent metabolites. aNCA uses
  this to set the `METABFL` flag.
- **`NCAwXRS`**: One or more columns containing NCA exclusion reason codes.
  Subjects flagged here are excluded from calculations.
- **`Grouping_Variables`**: Columns carried through to results and used for
  grouping in descriptive statistics (e.g., treatment arm, demographics).

### 3. Set up filters
Filters restrict the data before analysis. Each filter specifies a column and
the values to keep. Here we keep only the parent drug for serum specimens:
```{r filters, eval = FALSE}
applied_filters <- list(
  list(column = "PARAM",  value = c("DrugA")),
  list(column = "PCSPEC", value = c("SERUM"))
)
```
Pass `NULL` if no filters are needed.

### 4. Define NCA settings

#### 4.1 Partial interval calculations
Interval parameters compute AUC or average concentration over custom time
windows. Each row specifies the parameter code, start time and end time:
```{r int_parameters, eval = FALSE}
int_parameters <- data.frame(
  parameter = c("AUCINT", "CAVGINT"),
  start_auc = c(0, 0),
  end_auc   = c(12, 12)
)
```
This calculates for each dose AUCs from 0–12 h and average concentration from 0–12 h.

#### 4.2 Parameter selections
Select which NCA parameters to compute, grouped by study type. The names must
match the study types that PKNCA assigns based on route and dosing:
```{r parameters, eval = FALSE}
parameters_selected_per_study <- list(
  `Single Extravascular` = c(
    "auclast", "cmax", "tmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs", "aumclast"
  ),
  `Single IV Bolus` = c(
    "auclast", "cmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs"
  )
)
```

#### 4.3 Custom units
Override the default units for specific parameters. Each row maps a PPTESTCD to
the desired output unit:
```{r units, eval = FALSE}
units_table <- data.frame(
  PPTESTCD = c("CMAX", "TMAX", "AUCLAST"),
  PPSTRESU = c("ng/mL", "h", "ng*h/mL")
)
```
Pass `NULL` to use the units derived from the data.

#### 4.4 Flag rules
Flag rules define quality thresholds. Parameters that violate a checked rule are
flagged and excluded from descriptive statistics. Each rule has a checkbox
(`is.checked`) and a numeric `threshold`:
```{r flags, eval = FALSE}
flag_rules <- list(
  R2ADJ = list(is.checked = TRUE,  threshold = 0.7),
  R2    = list(is.checked = FALSE, threshold = 0.7),
  AUCPEO  = list(is.checked = TRUE,  threshold = 20),
  AUCPEP  = list(is.checked = TRUE,  threshold = 20),
  LAMZSPN = list(is.checked = TRUE,  threshold = 2)
)
```
- **R2ADJ ≥ 0.7**: Minimum adjusted R² for the terminal elimination regression.
- **AUCPEO ≤ 20**: Maximum % AUC extrapolated (observed).
- **AUCPEP ≤ 20**: Maximum % AUC extrapolated (predicted).
- **LAMZSPN ≥ 2**: Minimum half-life span ratio.

#### 4.5 Slope rules
Slope rules let you manually override which points PKNCA uses for the
terminal elimination (lambda-z) regression. This is useful when the automatic
best-fit selects an inappropriate range — for example, when a secondary
absorption peak or an outlier distorts the slope.

Each row targets one subject/profile combination:

- **`USUBJID`**: Subject identifier.
- **`ATPTREF`**: Dosing period (e.g., `"DOSE 1"`).
- **`DOSNOA`**: Dose number within the period (use `"*"` for all doses).
- **`TYPE`**: Either `"points"` (select individual time points) or `"range"`
  (select a start–end window).
- **`RANGE`**: The time points or window, e.g., `"4, 8, 12, 24"` for points
  or `"4-24"` for a range.
- **`REASON`**: Free-text justification for the override.

```{r slope_rules, eval = FALSE}
# Example: force lambda-z to use 4–24 h for subject S1-001 in DOSE 1
slope_rules <- data.frame(
  USUBJID = "S1-001",
  ATPTREF = "DOSE 1",
  DOSNOA  = "*",
  TYPE    = "range",
  RANGE   = "4-24",
  REASON  = "Exclude early absorption phase from terminal slope"
)

# If no manual adjustments are needed, pass an empty data frame:
# slope_rules <- data.frame(
#   USUBJID = character(0), ATPTREF = character(0),
#   DOSNOA  = character(0), TYPE    = character(0),
#   RANGE   = character(0), REASON  = character(0)
# )
```

#### 4.6 Extra variables to keep
These columns are carried through from the input data to the final results.
They always include `DOSEA`, `ATPTREF`, and `ROUTE`, plus whatever you selected
as `Grouping_Variables` in the mapping:
```{r extra_vars, eval = FALSE}
extra_vars_to_keep <- c(
  mapping$Grouping_Variables, "DOSEA", "ATPTREF", "ROUTE"
)
```

### 5. Create the PKNCA data object
`PKNCA_create_data_object()` applies the column mapping, filters, and duplicate
handling to produce a PKNCA-ready data object:
```{r create_obj, eval = FALSE}
pknca_obj <- adnca_data %>%
  PKNCA_create_data_object(
    mapping = mapping,
    applied_filters = applied_filters,
    time_duplicate_rows = NULL
  )
```
`time_duplicate_rows` is a character vector of column names used to identify
duplicate time points. Pass `NULL` to skip duplicate handling.

### 6. Configure the data object
`PKNCA_update_data_object()` sets the NCA method, selects the analyte/profile/
specimen, configures intervals, parameters, and units:
```{r update_obj, eval = FALSE}
pknca_obj <- pknca_obj %>%
  PKNCA_update_data_object(
    method = "lin up/log down",
    selected_analytes = "DrugA",
    selected_profile = "DOSE 1",
    selected_pcspec = "SERUM",
    start_impute = "yes",
    exclusion_list = list(),
    hl_adj_rules = slope_rules,
    keep_interval_cols = setdiff(extra_vars_to_keep, c("DOSEA", "ATPTREF", "ROUTE")),
    min_hl_points = 3,
    parameter_selections = parameters_selected_per_study,
    int_parameters = int_parameters,
    custom_units_table = units_table
  )
```
Key arguments:
- **`method`**: AUC calculation method. Options: `"linear"`, `"lin up/log down"`,
  `"linear up/log down"`.
- **`selected_analytes`**: Which PARAM value(s) to analyze.
- **`selected_profile`**: Which ATPTREF (dosing period) to analyze.
- **`selected_pcspec`**: Which specimen type to use.
- **`start_impute`**: Whether to impute C0 (`"yes"` or `"no"`).
- **`exclusion_list`**: Subject-level exclusions (list of USUBJID values).
- **`min_hl_points`**: Minimum number of points for lambda-z regression.
- **`keep_interval_cols`**: Extra columns to carry into PKNCA intervals (the
  grouping variables, minus the three that are always included).

### 7. Run NCA calculations
```{r run_nca, eval = FALSE}
pknca_res <- pknca_obj %>%
  # Run PKNCA and join subject/dose information to results
  PKNCA_calculate_nca(
    blq_rule = list(first = "keep", middle = "keep", last = "keep")
  ) %>%
  # Add bioavailability if requested (NULL = skip)
  add_f_to_pknca_results(NULL) %>%
  # Translate PKNCA parameter names to CDISC PPTESTCD codes
  mutate(PPTESTCD = translate_terms(PPTESTCD, "PKNCA", "PPTESTCD"))
```
- **`blq_rule`**: How to handle BLQ (below limit of quantification) values at
  different positions in the concentration-time profile. Options for each
  position: `"keep"`, `"0"`, `"loq/2"`.
- **`add_f_to_pknca_results()`**: Pass a bioavailability AUC type
  (e.g., `"f_aucinf.obs"`) to compute F, or `NULL` to skip.

### 8. Apply flag rules
Flag rules mark parameters that fail quality thresholds. Flagged parameters
get an exclusion reason and are dropped from descriptive statistics:
```{r flag, eval = FALSE}
pknca_res <- pknca_res %>%
  PKNCA_hl_rules_exclusion(
    rules = flag_rules %>%
      purrr::keep(\(x) x$is.checked) %>%
      purrr::map(\(x) x$threshold)
  )
```
This filters `flag_rules` to only the checked rules, extracts their thresholds,
and passes them to PKNCA's exclusion logic.

### 9. Calculate ratio parameters
The ratio table derives secondary parameters by dividing one NCA result by
another across groups. Here we compute a metabolite-to-parent Cmax ratio:
```{r ratios, eval = FALSE}
ratio_table <- data.frame(
  TestParameter = "CMAX",
  RefParameter  = "CMAX",
  RefGroups     = "PARAM: DrugA",
  TestGroups    = "(all other levels)",
  AggregateSubject = "no",
  AdjustingFactor  = 1,
  PPTESTCD = "MRCMAX"
)
pknca_res <- pknca_res %>%
  calculate_table_ratios(ratio_table)
```
Pass `NULL` or an empty data frame if no ratios are needed.

### 10. Filter to requested parameters
Remove any parameters that were computed by PKNCA but not explicitly selected
in `parameters_selected_per_study`:
```{r filter_params, eval = FALSE}
pknca_res <- pknca_res %>%
  remove_pp_not_requested()
```

### 11. Export CDISC datasets
`export_cdisc()` produces PP, ADPP, and ADNCA datasets from the NCA results.
Flag rules are formatted into CRITy/CRITyFL/PPSUMFL columns in ADPP:
```{r cdisc, eval = FALSE}
# Build flag rule messages for ADPP criterion columns
flag_operators <- c(
  R2ADJ = " < ", R2 = " < ", AUCPEO = " > ", AUCPEP = " > ", LAMZSPN = " < "
)
checked_flags <- purrr::keep(flag_rules, function(x) x$is.checked)
flag_rule_msgs <- if (length(checked_flags) > 0) {
  vapply(names(checked_flags), function(nm) {
    paste0(nm, flag_operators[nm], checked_flags[[nm]]$threshold)
  }, character(1), USE.NAMES = FALSE)
} else {
  NULL
}
cdisc_datasets <- pknca_res %>%
  export_cdisc(
    grouping_vars = extra_vars_to_keep,
    flag_rules = flag_rule_msgs
  )
```
`cdisc_datasets` is a named list with three data frames:
- **`cdisc_datasets$pp`**: PP domain — one row per parameter per subject.
- **`cdisc_datasets$adpp`**: ADPP — PP with ADaM-style flags and criterion
  columns (CRITy, CRITyFL, PPSUMFL, PPSUMRSN).
- **`cdisc_datasets$adnca`**: ADNCA — the full analysis dataset with results
  joined back to the concentration-time data.

### 12. Pivot results
`pivot_wider_pknca_results()` reshapes the long-format NCA results into a
wide table with one row per subject and one column per parameter — the same
view shown in the app's Results tab:
```{r pivot, eval = FALSE}
pivoted_results <- pivot_wider_pknca_results(
  myres = pknca_res,
  flag_rules = flag_rules,
  extra_vars_to_keep = extra_vars_to_keep
)
```
Flagged parameters (those excluded by flag rules) are marked in the output
and excluded from summary statistics.

## Path B: generate the same script from YAML with get_settings_code()
If you prefer managing configuration as YAML, you can generate the same runnable
R script with `get_settings_code()`.

### Example YAML (same analysis setup)
Save this as `example_settings.yaml`, and feel free to edit based on preferences:

```yaml
mapping:
  STUDYID: STUDYID
  USUBJID: USUBJID
  DOSEA: DOSEA
  DOSEU: DOSEU
  DOSETRT: DOSETRT
  PARAM: PARAM
  Metabolites: Metab-DrugA
  ARRLT: ARRLT
  NRRLT: NRRLT
  AFRLT: AFRLT
  NCAwXRS: [NCA1XRS, NCA2XRS]
  NFRLT: NFRLT
  PCSPEC: PCSPEC
  ROUTE: ROUTE
  TRTRINT: TRTRINT
  ADOSEDUR: ADOSEDUR
  Grouping_Variables: [TRT01A, RACE, SEX]
  RRLTU: RRLTU
  VOLUME: VOLUME
  VOLUMEU: VOLUMEU
  AVAL: AVAL
  AVALU: AVALU
  ATPTREF: ATPTREF

filters:
  - column: PARAM
    value: [DrugA]
  - column: PCSPEC
    value: [SERUM]

settings:
  method: lin up/log down
  analyte: DrugA
  profile: DOSE 1
  pcspec: SERUM
  start_impute: yes
  min_hl_points: 3
  int_parameters:
    parameter: [AUCINT, CAVGINT]
    start_auc: [0, 0]
    end_auc: [12, 12]
  ratio_table:
    TestParameter: [CMAX]
    RefParameter: [CMAX]
    RefGroups: ["PARAM: DrugA"]
    TestGroups: ["(all other levels)"]
    AggregateSubject: [no]
    AdjustingFactor: [1]
    PPTESTCD: [MRCMAX]
  units:
    PPTESTCD: [CMAX, TMAX, AUCLAST]
    PPSTRESU: [ng/mL, h, ng*h/mL]

slope_rules:
  USUBJID: []
  ATPTREF: []
  DOSNOA: []
  TYPE: []
  RANGE: []
  REASON: []

time_duplicate_keys: null
```

### Generate the script from YAML
```{r yaml-to-script, eval = FALSE}
library(aNCA)

get_settings_code(
  settings_file_path = "example_settings.yaml",
  data_path = "adnca_example.csv",
  output_path = "settings_code.R"
)
```

This writes a runnable script to `settings_code.R` using your YAML settings (check the [Full script](#full-script) section below). It will have some comments explaining each step, but for more details refer back to [Path A: run the full script directly](#path-a-run-the-full-script-directly).

## Path C: generate from the aNCA app
If you prefer a guided UI workflow, it is very simple:

```{r run-app, eval = FALSE}
aNCA::run_app()
```

From the app, configure mapping, filters, and NCA settings, then export pressing the `Save` button:
- Settings YAML file
- Reproducible R script
- All the generated outputs of your interest

For a YAML-driven reproducible workflow after app export, see [Path B: generate the same script from YAML with get_settings_code()](#path-b-generate-the-same-script-from-yaml-with-get_settings_code).


## Full script
Below is the complete script assembled from the steps above. This is what the
app generates when you export settings — with all placeholders filled in.

<details>
<summary>Show full reproducible script</summary>

```{r full_script, eval = FALSE}
# Load the package
library(aNCA)
library(dplyr)
# 1. Load data
adnca_data <- adnca_example
# 2. Column mapping
mapping <- list(
  STUDYID = "STUDYID",
  USUBJID = "USUBJID",
  DOSEA   = "DOSEA",
  DOSEU   = "DOSEU",
  DOSETRT = "DOSETRT",
  PARAM   = "PARAM",
  Metabolites = "Metab-DrugA",
  ARRLT   = "ARRLT",
  NRRLT   = "NRRLT",
  AFRLT   = "AFRLT",
  NCAwXRS = c("NCA1XRS", "NCA2XRS"),
  NFRLT   = "NFRLT",
  PCSPEC  = "PCSPEC",
  ROUTE   = "ROUTE",
  TRTRINT = "TRTRINT",
  ADOSEDUR = "ADOSEDUR",
  Grouping_Variables = c("TRT01A", "RACE", "SEX"),
  RRLTU   = "RRLTU",
  VOLUME  = "VOLUME",
  VOLUMEU = "VOLUMEU",
  AVAL    = "AVAL",
  AVALU   = "AVALU",
  ATPTREF = "ATPTREF"
)
# 3. Filters
applied_filters <- list(
  list(column = "PARAM",  value = c("DrugA")),
  list(column = "PCSPEC", value = c("SERUM"))
)
# 4. NCA settings
int_parameters <- data.frame(
  parameter = c("AUCINT", "CAVGINT"),
  start_auc = c(0, 0),
  end_auc   = c(12, 12)
)
parameters_selected_per_study <- list(
  `Single Extravascular` = c(
    "auclast", "cmax", "tmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs", "aumclast"
  ),
  `Single IV Bolus` = c(
    "auclast", "cmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs"
  )
)
units_table <- data.frame(
  PPTESTCD = c("CMAX", "TMAX", "AUCLAST"),
  PPSTRESU = c("ng/mL", "h", "ng*h/mL")
)
flag_rules <- list(
  R2ADJ   = list(is.checked = TRUE,  threshold = 0.7),
  R2      = list(is.checked = FALSE, threshold = 0.7),
  AUCPEO  = list(is.checked = TRUE,  threshold = 20),
  AUCPEP  = list(is.checked = TRUE,  threshold = 20),
  LAMZSPN = list(is.checked = TRUE,  threshold = 2)
)
# Slope rules: manual lambda-z overrides (empty = no overrides)
slope_rules <- data.frame(
  USUBJID = "S1-001",
  ATPTREF = "DOSE 1",
  DOSNOA  = "*",
  TYPE    = "range",
  RANGE   = "4-24",
  REASON  = "Exclude early absorption phase from terminal slope"
)
extra_vars_to_keep <- c(
  mapping$Grouping_Variables, "DOSEA", "ATPTREF", "ROUTE"
)
ratio_table <- data.frame(
  TestParameter = "CMAX",
  RefParameter  = "CMAX",
  RefGroups     = "PARAM: DrugA",
  TestGroups    = "(all other levels)",
  AggregateSubject = "no",
  AdjustingFactor  = 1,
  PPTESTCD = "MRCMAX"
)
# 5. Create PKNCA data object
pknca_obj <- adnca_data %>%
  PKNCA_create_data_object(
    mapping = mapping,
    applied_filters = applied_filters,
    time_duplicate_rows = NULL
  )
# 6. Configure intervals, parameters, and units
pknca_obj <- pknca_obj %>%
  PKNCA_update_data_object(
    method = "lin up/log down",
    selected_analytes = "DrugA",
    selected_profile = "DOSE 1",
    selected_pcspec = "SERUM",
    start_impute = "yes",
    exclusion_list = list(),
    hl_adj_rules = slope_rules,
    keep_interval_cols = setdiff(extra_vars_to_keep, c("DOSEA", "ATPTREF", "ROUTE")),
    min_hl_points = 3,
    parameter_selections = parameters_selected_per_study,
    int_parameters = int_parameters,
    custom_units_table = units_table
  )
# 7. Run NCA
pknca_res <- pknca_obj %>%
  PKNCA_calculate_nca(
    blq_rule = list(first = "keep", middle = "keep", last = "keep")
  ) %>%
  add_f_to_pknca_results(NULL) %>%
  mutate(PPTESTCD = translate_terms(PPTESTCD, "PKNCA", "PPTESTCD"))
# 8. Apply flag rules
pknca_res <- pknca_res %>%
  PKNCA_hl_rules_exclusion(
    rules = flag_rules %>%
      purrr::keep(\(x) x$is.checked) %>%
      purrr::map(\(x) x$threshold)
  )
# 9. Calculate ratios
pknca_res <- pknca_res %>%
  calculate_table_ratios(ratio_table)
# 10. Filter to requested parameters
pknca_res <- pknca_res %>%
  remove_pp_not_requested()
# 11. Export CDISC datasets
flag_operators <- c(
  R2ADJ = " < ", R2 = " < ", AUCPEO = " > ", AUCPEP = " > ", LAMZSPN = " < "
)
checked_flags <- purrr::keep(flag_rules, function(x) x$is.checked)
flag_rule_msgs <- if (length(checked_flags) > 0) {
  vapply(names(checked_flags), function(nm) {
    paste0(nm, flag_operators[nm], checked_flags[[nm]]$threshold)
  }, character(1), USE.NAMES = FALSE)
} else {
  NULL
}
cdisc_datasets <- pknca_res %>%
  export_cdisc(
    grouping_vars = extra_vars_to_keep,
    flag_rules = flag_rule_msgs
  )
# 12. Pivot results
pivoted_results <- pivot_wider_pknca_results(
  myres = pknca_res,
  flag_rules = flag_rules,
  extra_vars_to_keep = extra_vars_to_keep
)
```
</details>
